A Responsible Machine Learning Workflow\\\normalsize{With Focus on Interpretable Models, Post-hoc Explanation, and Discrimination Testing}

### 2020-02-02 article-information-2019.tex
a perhaps not broken structure that adheres to the suggested 16-page limit

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/us/en/services/consulting/analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}.} Although responsible AI is today a somewhat broad and amorphous notion, at least one aspect is becoming clear. ML models, a common application of AI, can present serious risks. ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{\textit{When a Computer Program Keeps You in Jail}}.} ML models can perpetuate and exacerbate discrimination \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}.  ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}, \cite{shokri2019privacy}. This manuscript makes no claim that these interdependent issues of opaqueness, discrimination, privacy harms, and security vulnerabilities in ML have been resolved, even as singular entities, and much less as complex intersectional phenomena. However, Sections \ref{sec:m_and_m}, \ref{sec:res}, and \ref{sec:disc} do propose some specific technical countermeasures, in the form interpretable models, post-hoc explanation, and DI and discrimination testing implemented in widely available Python tools, that responsible practitioners can use to address a subset of these vexing problems.\footnote{This text and associated software are not, and should not be construed as, legal advice or requirements for regulatory compliance.}\textsuperscript{,}\footnote{In the United States (US), interpretable models, explanations, DI testing, and the model documentation they enable may be required under the Civil Rights Acts of 1964 and 1991, the Americans with Disabilities Act, the Genetic Information Nondiscrimination Act, the Health Insurance Portability and Accountability Act, the Equal Credit Opportunity Act (ECOA), the Fair Credit Reporting Act (FCRA), the Fair Housing Act, Federal Reserve SR 11-7, and the European Union (EU) Greater Data Privacy Regulation (GDPR) Article 22 \cite{ff_interpretability}.}

@@

Two $g^\text{XNN}$ models are trained by mini-batch stochastic gradient descent (SGD) on the simulated data and mortgage data. Each $g^\text{XNN}$ is assessed in 5 training folds and in a test data partition. $L_1$ regularization is applied to all network weights to induce a sparse and interpretable model, where each $n_k$ and corresponding $\gamma_k$ are ideally associated with an important $X_j$ or combination thereof. The $g^\text{XNN}$ models appear highly sensitive to weight initialization and batch size. Be aware that $g^{XNN}$ model architectures may require manual and judicious feature selection due to long training times. For more details regarding $g^\text{XNN}$ training, see the software resources in Subsection \ref{ssec:soft} and Appendices  \ref{a_ssec:not} and \ref{a_ssec:xnn}.

@@

Results are laid out for the simulated and mortgage datasets. Accuracy is compared for unconstrained, less interpretable $g^{\text{GBM}}$ and $g^{\text{ANN}}$ models and constrained, more interpretable $g^{\text{MGBM}}$ and $g^{\text{XNN}}$ models. Then, for the  $g^{\text{MGBM}}$ and $g^{\text{XNN}}$ models, intrinsic interpretability, post-hoc explanation, and discrimination testing results are presented.

@@

As model training and assessment on the simulated data is a rough validation exercise meant to showcase expected results on data with known characteristics, and given that most of the techniques in the proposed workflow are already used widely or have been validated elsewhere, reporting of simulated data results in the main text will focus mostly on challenging the perceived \textit{accuracy-interpretability tradeoff} for the considered model architectures and on the more novel XNN interpretability results. The bulk of the post-hoc explanation and discrimination testing results for the simulated are left to Appendix \ref{a_sec:res_sim}.

@@

Model fit is roughly uniform for $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the simulated test data. Interpretability and explainability benefits of the constrained models appear to come at little cost to overall model performance, or in the case of $g^\text{ANN}$ and $g^\text{XNN}$, no cost at all. $g^\text{XNN}$ actually shows slightly better fit than $g^\text{ANN}$ across accuracy, area under the curve (AUC), logloss, and root mean squared error (RMSE). Accuracy is measured at the best F1 threshold for each model. Table \ref{tab:sim_acc} presents a variety of fit metrics for the $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the simulated test data. $g^\text{GBM}$ exhibits the best performance, but all models give relatively similar fit results. Interpretability and explainability benefits of the constrained models appear to come at little cost to overall model performance, or in the case of $g^\text{ANN}$ and $g^\text{XNN}$, no cost at all. $g^\text{XNN}$ actually shows slightly better fit than $g^\text{ANN}$ across accuracy, area under the curve (AUC), logloss, and root mean squared error (RMSE). Measures that require a probability cutoff are taken at the best F1 threshold for each model.

@@

% add error metrics

\subsubsection{$g^\text{XNN}$ Interpretability Results}

For $g^\text{XNN}$, inherent interpretability manifests as plots of sparse $\gamma_k$ output layer weights, $n_k$ subnetwork ridge functions, and sparse $\beta_j$ weights in the projection layer. Figure \ref{fig:sim_xnn_glob_ridge} provides detailed insights into $g^\text{XNN}$. \ref{fig:sim_xnn_glob_ridge} \textbf{A} displays the sparse $\gamma_k$ weights of the output layer, where only $n_k$ subnetworks with $k \in \{1, 4, 7, 8, 9\}$ are associated with large magnitude weights. The $n_k$ subnetwork ridge functions appear in \ref{fig:sim_xnn_glob_ridge} \textbf{B} as simplistic but distinctive functional forms. Color-coding between \ref{fig:sim_xnn_glob_ridge} \textbf{A} and \ref{fig:sim_xnn_glob_ridge} \textbf{B} visually reinforces the direct feed-forward relationship between the $n_k$ subnetworks and the $\gamma_k$ weights of the output layer.

@@

% update figure

\noindent $n_k$ subnetworks are plotted across the output values of their associated $\sum_j\beta_{k,j}x_j$ projection layer hidden units, and color-coding between \ref{fig:sim_xnn_glob_ridge} \textbf{B} and \ref{fig:sim_xnn_glob_ridge} \textbf{C} link the $\beta_j$ weights to their $n_k$ subnetworks. Most of the heavily utilized $n_k$ subnetworks have sparse weights in their $\sum_j\beta_{k,j}x_j$ projection layer hidden units. In particular, subnetwork $n_1$ appears to be almost solely a function of $X_{\text{Friedman}, 3}$ and appears to exhibit the expected quadratic behavior for $X_{\text{Friedman}, 3}$.  Subnetworks $n_7$, $n_8$, and $n_9$ appear to be most associated with the globally important $X_{\text{Friedman}, 1}$ and $X_{\text{Friedman}, 2}$ features, likely betraying the effort required for $g^{\text{XNN}}$ to model the nonlinear $\text{sin}()$ function of the $X_{\text{Friedman}, 1}$ and $X_{\text{Friedman}, 2}$ product, and $n_7$, $n_8$, and $n_9$ do appear to display some noticeable sinusoidal characteristics.  Conversely, subnetwork $n_4$ seems to be a linear combination of all the original input $X_j$ features, but does weigh the linear $X_{\text{Friedman},4}$ and $X_{\text{Friedman},5}$ terms roughly in the correct two-to-one ratio. As a whole, Figure \ref{fig:sim_xnn_glob_ridge} \textbf{A}, \textbf{B}, and \textbf{C} exhibit evidence that $g^\text{XNN}(\mathbf{X})$ has learned about the simulated function in Equation \ref{eq:sim} and the displayed information should help practitioners understand which original input $X_j$ features are weighed heavily in each $n_k$ subnetwork, and which $n_k$ subnetworks have a strong influence on $g^\text{XNN}(\mathbf{X})$.

@@

Results for the mortgage data are presented in Subsections \ref{ssec:c_v_uc_mort} -- \ref{ssec:dis_mort} to showcase the example workflow.  $g^\text{ANN}$ and $g^\text{XNN}$ outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data, but as in Subsection \ref{ssec: c_v_uc_sim} the constrained variants of both model architectures do not show large differences in model performance with respect to unconstrained variants. Assuming that small fit differences on static test data do not outweigh the need for intrinsic model interpretability and reliable post-hoc explainability in high-stakes, human-centered, or regulated applications, only $g^\text{MGBM}$ and $g^\text{XNN}$ interpretability, post-hoc explainability, and discrimination testing results are presented.

@@

\begin{table}[H]
	\caption{Fit measures for $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the mortgage test data. Arrows indicate the direction of improvement for each measure.}
\centering
\begin{tabular}{ccccccccc}
	\toprule
	\textbf{Model} & 
	\href{https://scikit-learn.org/stable/modules/model_evaluation.html#accuracy-score}{\textbf{Accuracy}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Receiver_operating_characteristic#Area_under_the_curve}{\textbf{AUC}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/F1_score}{\textbf{F1}}~$\uparrow$ & 
	\href{https://scikit-learn.org/stable/modules/model_evaluation.html#log-loss}{\textbf{Logloss}}~$\downarrow$ & \href{https://en.wikipedia.org/wiki/Matthews_correlation_coefficient}{\textbf{MCC}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Root-mean-square_deviation}{\textbf{RMSE}}~$\downarrow$ & \href{https://en.wikipedia.org/wiki/Sensitivity_and_specificity}{\textbf{Sensitivity}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Sensitivity_and_specificity}{\textbf{Specificity}}~$\uparrow$ \\
	\midrule	
$g^\text{GBM}$		& 0.795	& 0.828	& & 0.252 &	& 0.276 & & \\
$g^\text{MGBM}$		& 0.765	& 0.814	& & 0.259 &	& 0.278 & & \\
$g^\text{ANN}$		& 0.865	& 0.871	& 0.474 & 0.231 &	0.418 & 0.262 & 0.624 & \\
$g^\text{XNN}$		& 0.869	& 0.868	& 0.468 & 0.233 &	0.409 & 0.263 & 0.593 & \\
\bottomrule
\label{tab:mort_acc}
\end{tabular}
\end{table}

@@

@@

Global Shapley feature importance for $g^\text{MGBM}$ on the mortgage test data is reported in Figure \ref{fig:mort_mgbm_glob}. $g^\text{MGBM}$ places high importance on LTV ratio, perhaps too high, and also weighs DTI ratio, property value, loan amount, and introductory rate period heavily in many of its predictions.

\begin{figure}[H]
\centering
\includegraphics[width=9cm]{img/mort_mgbm_glob.png}
\caption{Global mean absolute Tree SHAP feature importance for $g^\text{MGBM}(\mathbf{X})$ on the mortgage test data.}
\label{fig:mort_mgbm_glob}
\end{figure}

\noindent The potential over-emphasis of LTV ratio, and the de-emphasis of income, likely an important feature from a business perspective, and the encoded no introductory rate period flag feature may contribute to the decreased performance of $g^\text{MGBM}$ as compared to $g^\text{XNN}$.

Domain knowledge was used to positively constrain DTI ratio and LTV ratio and to negatively constrain income and the loan term flag under $g^\text{MGBM}$. The monotonicity constraints for DTI ratio and LTV ratio are confirmed for $g^\text{MGBM}(\mathbf{X})$ on the mortgage test data in Figure \ref{fig:mort_mgbm_glob_pdp_ice}. Both DTI ratio and LTV ratio display positive monotonic behavior at all selected percentiles for ICE and on average with PD. Because PD curves generally follow the patterns of the ICE curves for both features, it's also likely that no strong interactions are at play for DTI ratio and LTV ratio under $g^\text{MGBM}$. Of course, the monotonicity constraints themselves can dampen the effects of non-monotonic interactions under $g^\text{MGBM}$, even if they do exist in the data, and perhaps due to less noise in the mortgage data, this rigidity could also play a role in the performance differences between $g^\text{MGBM}$ and $g^\text{XNN}$ in the mortgage data. DTI ratio and LTV ratio also appear to have sparse regions in their univariate distributions. The monotonicity constraints likely play to the advantage of $g^\text{MGBM}$ in this regard, as $g^\text{MGBM}$ appears to carry reasonable predictions learned from populous domains into the sparse domains of both features.

\begin{figure}[H]
\centering
\includegraphics[width=16.5cm]{img/mort_mgbm_glob_pdp_ice.png}
\caption{PD, ICE for 10 individuals across selected percentiles of $g^\text{MGBM}(\mathbf{X})$, and histograms for the three most important input features of $g^\text{MGBM}$ on the mortgage test data.}
\label{fig:mort_mgbm_glob_pdp_ice}
\end{figure}

@@

In Figure \ref{fig:mort_mgbm_loc}, local Tree SHAP values are displayed for selected individuals at the 10\textsuperscript{th}, 50\textsuperscript{th}, and 90\textsuperscript{th} percentiles of $g^{\text{MGBM}}(\mathbf{X})$ in the mortgage test data. The selected individuals show an expected progression of mostly negative Shapley values at the 10\textsuperscript{th} percentile, a mixture of positive and negative Shapley values at the 50\textsuperscript{th} percentile, mostly positive Shapley values the 90\textsuperscript{th} percentile, and with globally important features driving most local model decisions.

\begin{figure}[H]
\centering
\includegraphics[width=16cm]{img/mort_mgbm_loc.png}
\caption{Tree SHAP values for three individuals across selected percentiles of $g^\text{MGBM}(\mathbf{X})$ for the mortgage test data.}
\label{fig:mort_mgbm_loc}
\end{figure}

\noindent Deeper significance for Figure \ref{fig:mort_mgbm_loc} lies in the ability of Tree SHAP to \textit{accurately} summarize any single $g^\text{MGBM}(\mathbf{x})$ prediction in this manner, which is generally important for enabling logical appeal or override of ML-based decisions, and is specifically important in the context of lending, where applicable regulations often require lenders to provide consumer-specific reasons for denying credit to an individual. In the US, applicable regulations are typically ECOA and FCRA, and the consumer-specific reasons are commonly known as adverse actions codes.

Figure \ref{fig:mort_xnn_glob} displays global feature importance for $g^\text{XNN}$ on the mortgage test data. $g^\text{XNN}$ distributes importance more evenly across business drivers and puts stronger emphasis on the no introductory rate period flag feature than does $g^\text{MGBM}$. Like $g^\text{MGBM}$, $g^\text{XNN}$ puts little emphasis on the other flag features.

\begin{figure}[H]
\centering
\includegraphics[width=9cm]{img/mort_xnn_glob.png}
\caption{Global mean absolute Deep SHAP feature importance for $g^\text{XNN}(\mathbf{X})$ on the mortgage test data.}
\label{fig:mort_xnn_glob}
\end{figure}

@@

Figure \ref{fig:blueprint} displays a holistic approach to ML model training, assessment, deployment meant to decrease discrimination, inaccuracy, privacy, and security risks for high-stakes, human-centered, or regulated ML applications.\footnote{See: \href{https://github.com/jphall663/hc_ml}{Toward Responsible Machine Learning} for details regarding Figure \ref{fig:blueprint}.}

@@

\noindent See: \href{https://github.com/jphall663/awesome-machine-learning-interpretability}{Awesome Machine Learning Interpretability} for a longer, community-curated metalist of related software packages and resources.

@@

The heightened interest in exploring and fixing algorithmic bias has led to a significant body of work focused in using ML to diminish discrimination in model outcomes \cite{friedler2019comparative}. Broadly, these can be defined into two groups: more traditional methods that mitigate discrimination by searching across possible algorithmic and feature specifications, and more recently developed approaches that change the algorithms or input data themselves in order to mitigate disparities. Many approaches that have been developed in the last five to seven years focus on altering the algorithm itself, preprocessing the data, or post-processing the predictions in order to diminish class-control correlations or dependencies. For comments on why these techniques could result in regulatory non-compliance in certain cases, see Appendix \ref{a_sec:comp}.

>>>>

Responsible artificial intelligence (AI) has been variously conceptualized as AI-based products or projects that use transparent technical mechanisms, that create appealable decisions or outcomes, that perform reliably and in a trustworthy manner over time, that exhibit minimal social discrimination, and that are designed by humans with diverse experiences, both in terms of demographics and professional backgrounds.\footnote{See: \href{https://ec.europa.eu/jrc/communities/sites/jrccties/files/03_dignum_v.pdf}{Responsible Artificial Intelligence}, \href{https://www.accenture.com/_acnmedia/pdf-92/accenture-afs-responsible-ai.pdf}{\textit{Responsible AI: A Framework for Building Trust in Your AI Solutions}}, \href{https://www.pwc.com/us/en/services/consulting/analytics/artificial-intelligence/what-is-responsible-ai.html}{PwC's Responsible AI}, \href{https://ai.google/responsibilities/responsible-ai-practices/}{Responsible AI Practices}.} Although responsible AI is today a somewhat broad and amorphous notion, at least one aspect is becoming clear. ML models, a common application of AI, can present serious risks. ML models can be inaccurate and unappealable black-boxes, even with the application of newer post-hoc explanation techniques \cite{please_stop}.\footnote{See: \href{https://www.nytimes.com/2017/06/13/opinion/how-computers-are-harming-criminal-justice.html}{\textit{When a Computer Program Keeps You in Jail}}.} ML models can perpetuate and exacerbate discrimination \cite{feldman2015certifying}, \cite{dwork2012fairness}, \cite{gender_shades}.  ML models can be hacked, resulting in manipulated model outcomes or the exposure of proprietary intellectual property or sensitive training data \cite{security_of_ml}, \cite{model_stealing}, \cite{membership_inference}, \cite{shokri2019privacy}. This manuscript makes no claim that these interdependent issues of opaqueness, discrimination, privacy harms, and security vulnerabilities in ML have been resolved, even as singular entities, and much less as complex intersectional phenomena. However, Sections \ref{sec:m_and_m}, \ref{sec:res}, and \ref{sec:disc} do propose some specific technical countermeasures, mostly in the form interpretable models, post-hoc explanation, and DI and discrimination testing implemented in widely available Python tools, that responsible practitioners can use to address a subset of these vexing problems.\footnote{This text and associated software are not, and should not be construed as, legal advice or requirements for regulatory compliance.}\textsuperscript{,}\footnote{In the United States (US), interpretable models, explanations, DI testing, and the model documentation they enable may be required under the Civil Rights Acts of 1964 and 1991, the Americans with Disabilities Act, the Genetic Information Nondiscrimination Act, the Health Insurance Portability and Accountability Act, the Equal Credit Opportunity Act (ECOA), the Fair Credit Reporting Act (FCRA), the Fair Housing Act, Federal Reserve SR 11-7, and the European Union (EU) Greater Data Privacy Regulation (GDPR) Article 22 \cite{ff_interpretability}.}

@@

Two $g^\text{XNN}$ models are trained by mini-batch stochastic gradient descent (SGD) on the simulated data and mortgage data. Each $g^\text{XNN}$ is assessed in 5 training folds and in a test data partition. $L_1$ regularization is applied to network weights to induce a sparse and interpretable model, where each $n_k$ and corresponding $\gamma_k$ are ideally associated with an important $X_j$ or combination thereof. The $g^\text{XNN}$ models appear highly sensitive to weight initialization and batch size. Be aware that $g^{XNN}$ model architectures may require manual and judicious feature selection due to long training times. For more details regarding $g^\text{XNN}$ training, see the software resources in Subsection \ref{ssec:soft} and Appendices  \ref{a_ssec:not} and \ref{a_ssec:xnn}.

@@

Results are laid out for the simulated and mortgage datasets. Accuracy is compared for unconstrained, less interpretable $g^{\text{GBM}}$ and $g^{\text{ANN}}$ models and constrained, more interpretable $g^{\text{MGBM}}$ and $g^{\text{XNN}}$ models. Then, for the  $g^{\text{MGBM}}$ and $g^{\text{XNN}}$ models, intrinsic interpretability, post-hoc explanation, and discrimination testing results are explored.

@@

Fit comparisons between unconstrained and constrained models and XNN interpretabililty results are discussed in Subsections \ref{ssec:c_v_uc_sim} and \ref{ssec:xnn_sim}. As model training and assessment on the simulated data is a rough validation exercise meant to showcase expected results on data with known characteristics, and given that most of the techniques in the proposed workflow are already used widely or have been validated elsewhere, reporting of simulated data results in the main text will focus mostly on challenging the perceived \textit{accuracy-interpretability tradeoff} for the considered model architectures and on the more novel XNN interpretability results. The bulk of the post-hoc explanation and discrimination testing results for the simulated data are left to Appendix \ref{a_sec:res_sim}.

@@

Table \ref{tab:sim_acc} presents a variety of fit metrics for the $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the simulated test data. $g^\text{GBM}$ exhibits the best performance, but all models exhibit only a small range fit results. Interpretability and explainability benefits of the constrained models appear to come at little cost to overall model performance, or in the case of $g^\text{ANN}$ and $g^\text{XNN}$, no cost at all. $g^\text{XNN}$ actually shows slightly better fit than $g^\text{ANN}$ across all recorded fit metrics. Fit measures that require a probability cutoff are taken at the best F1 threshold for each model.

@@

% TODO: add error metrics, confirm statements about results
\subsubsection{$g^\text{XNN}$ Interpretability Results}\label{ssec:xnn_sim}

For $g^\text{XNN}$, inherent interpretability manifests as plots of sparse $\gamma_k$ output layer weights, $n_k$ subnetwork ridge functions, and sparse $\beta_j$ weights in the bottom projection layer. Figure \ref{fig:sim_xnn_glob_ridge} provides detailed insights into $g^\text{XNN}$. \ref{fig:sim_xnn_glob_ridge} \textbf{A} displays the sparse $\gamma_k$ weights of the output layer, where only $n_k$ subnetworks with $k \in \{1, 4, 7, 8, 9\}$ are associated with large magnitude weights. The $n_k$ subnetwork ridge functions appear in \ref{fig:sim_xnn_glob_ridge} \textbf{B} as simplistic but distinctive functional forms. Color-coding between \ref{fig:sim_xnn_glob_ridge} \textbf{A} and \ref{fig:sim_xnn_glob_ridge} \textbf{B} visually reinforces the direct feed-forward relationship between the $n_k$ subnetworks and the $\gamma_k$ weights of the output layer.

@@

% TODO: update figure: connections, feed-forward direction, make subfigure reference match discrimination measure table

\noindent $n_k$ subnetworks are plotted across the output values of their associated $\sum_j\beta_{k,j}x_j$ projection layer hidden units, and color-coding between \ref{fig:sim_xnn_glob_ridge} \textbf{B} and \ref{fig:sim_xnn_glob_ridge} \textbf{C} link the $\beta_j$ weights to their $n_k$ subnetworks. Most of the heavily utilized $n_k$ subnetworks have sparse weights in their $\sum_j\beta_{k,j}x_j$ projection layer hidden units. In particular, subnetwork $n_1$ appears to be almost solely a function of $X_{\text{Friedman}, 3}$ and appears to exhibit the expected quadratic behavior for $X_{\text{Friedman}, 3}$.  Subnetworks $n_7$, $n_8$, and $n_9$ appear to be most associated with the globally important $X_{\text{Friedman}, 1}$ and $X_{\text{Friedman}, 2}$ features, likely betraying the effort required for $g^{\text{XNN}}$ to model the nonlinear $\text{sin}()$ function of the $X_{\text{Friedman}, 1}$ and $X_{\text{Friedman}, 2}$ product, and these subnetworks, especially $n_7$ and $n_8$, appear to display some noticeable sinusoidal characteristics. Subnetwork $n_4$ seems to be a linear combination of all the original input $X_j$ features, but does weigh the linear $X_{\text{Friedman},4}$ and $X_{\text{Friedman},5}$ terms roughly in the correct two-to-one ratio. As a whole, Figure \ref{fig:sim_xnn_glob_ridge} \textbf{A}, \textbf{B}, and \textbf{C} exhibit evidence that $g^\text{XNN}(\mathbf{X})$ has learned about the signal-generating function in Equation \ref{eq:sim} and the displayed information should help practitioners understand which original input $X_j$ features are weighed heavily in each $n_k$ subnetwork, and which $n_k$ subnetworks have a strong influence on $g^\text{XNN}(\mathbf{X})$ output.

@@

Results for the mortgage data are presented in Subsections \ref{ssec:c_v_uc_mort} -- \ref{ssec:dis_mort} to showcase the example workflow.  $g^\text{ANN}$ and $g^\text{XNN}$ outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data, but as in Subsection \ref{ssec:c_v_uc_sim} the constrained variants of both model architectures do not show large differences in model fit with respect to unconstrained variants. Assuming that in high-stakes applications small fit differences on static test data do not outweigh the need for enhanced model debugging facilitated by high interpretability, only $g^\text{MGBM}$ and $g^\text{XNN}$ interpretability, post-hoc explainability, and discrimination testing results are presented.

@@

\begin{table}[htb]
	\caption{Fit measures for $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the mortgage test data. Arrows indicate the direction of improvement for each measure.}
\centering
\begin{tabular}{ccccccccc}
	\toprule
	\textbf{Model} & 
	\href{https://scikit-learn.org/stable/modules/model_evaluation.html#accuracy-score}{\textbf{Accuracy}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Receiver_operating_characteristic#Area_under_the_curve}{\textbf{AUC}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/F1_score}{\textbf{F1}}~$\uparrow$ & 
	\href{https://scikit-learn.org/stable/modules/model_evaluation.html#log-loss}{\textbf{Logloss}}~$\downarrow$ & \href{https://en.wikipedia.org/wiki/Matthews_correlation_coefficient}{\textbf{MCC}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Root-mean-square_deviation}{\textbf{RMSE}}~$\downarrow$ & \href{https://en.wikipedia.org/wiki/Sensitivity_and_specificity}{\textbf{Sensitivity}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Sensitivity_and_specificity}{\textbf{Specificity}}~$\uparrow$ \\
	\midrule	
$g^\text{GBM}$		& 0.795	& 0.828	& & 0.252 &	& 0.276 & & \\
$g^\text{MGBM}$		& 0.765	& 0.814	& & 0.259 &	& 0.278 & & \\
$g^\text{ANN}$		& 0.865	& 0.871	& 0.474 & 0.231 &	0.418 & 0.262 & 0.624 & \\
$g^\text{XNN}$		& 0.869	& 0.868	& 0.468 & 0.233 &	0.409 & 0.263 & 0.593 & \\
\bottomrule
\label{tab:mort_acc}
\end{tabular}
\end{table}

@@

% TODO: add error metrics, confirm statements about results

@@

For $g^\text{MGBM}$, intrinsic interpretability is evaluated with PD and ICE plots of mostly monotonic prediction behavior for several important $X_j$, and post-hoc Shapley explanation analysis is used to create global and local feature importance. Global Shapley feature importance for $g^\text{MGBM}$ on the mortgage test data is reported in Figure \ref{fig:mort_mgbm_glob}. $g^\text{MGBM}$ places high importance on loan-to-value (LTV) ratio, perhaps too high, and also weighs debt-to-income (DTI) ratio, property value, loan amount, and introductory rate period heavily in many of its predictions. Tree SHAP values are reported in the margin space, prior to the application of the logit link function, and the reported numeric values can be interpreted as the mean absolute impact of each $X_j$ on $g^\text{MGBM}(\mathbf{X})$ in the mortgage test data in the $g^\text{MGBM}$ margin space.

\begin{figure}[htb]
\centering
\includegraphics[width=9cm]{img/mort_mgbm_glob.png}
\caption{Global mean absolute Tree SHAP feature importance for $g^\text{MGBM}(\mathbf{X})$ on the mortgage test data.}
\label{fig:mort_mgbm_glob}
\end{figure}

\noindent The potential over-emphasis of LTV ratio, and the de-emphasis of income, likely an important feature from a business perspective, and the encoded no introductory rate period flag feature may also contribute to the decreased performance of $g^\text{MGBM}$ as compared to $g^\text{XNN}$.

Domain knowledge was used to positively constrain DTI ratio and LTV ratio and to negatively constrain income and the loan term flag under $g^\text{MGBM}$. The monotonicity constraints for DTI ratio and LTV ratio are confirmed for $g^\text{MGBM}(\mathbf{X})$ on the mortgage test data in Figure \ref{fig:mort_mgbm_glob_pdp_ice}. Both DTI ratio and LTV ratio display positive monotonic behavior at all selected percentiles of $g^\text{MGBM}(\mathbf{X})$ for ICE and on average with PD. Because PD curves generally follow the patterns of the ICE curves for both features, it's also likely that no strong interactions are at play for DTI ratio and LTV ratio under $g^\text{MGBM}$. Of course, the monotonicity constraints themselves can dampen the effects of non-monotonic interactions under $g^\text{MGBM}$, even if they do exist in the data, and perhaps due to non-synthetic noise in the mortgage data, this rigidity could also play a role in the performance differences between $g^\text{MGBM}$ and $g^\text{XNN}$ in the mortgage data not observed in Table \ref{tab:sim_acc} for the simulated data.

PD and ICE are displayed with a histogram in Section \ref{sec:res} to highlight any sparse regions in an input feature's domain. Because most ML models will always issue a prediction on any datum with a correct schema, it's crucial to consider whether a given model learned enough about an observation to make an accurate prediction. Viewing PD and ICE along with a histogram is a convenient method to visually assess whether a prediction is reasonable and based on sufficient training data. DTI ratio and LTV ratio do appear to have sparse regions in their univariate distributions. The monotonicity constraints likely play to the advantage of $g^\text{MGBM}$ in this regard, as $g^\text{MGBM}$ appears to carry reasonable predictions learned from populous domains into the sparse domains of both features.

\begin{figure}[htb]
\hskip-10pt\includegraphics[width=16.5cm]{img/mort_mgbm_glob_pdp_ice.png}
\caption{PD, ICE for 10 individuals across selected percentiles of $g^\text{MGBM}(\mathbf{X})$, and histograms for the three most important input features of $g^\text{MGBM}$ on the mortgage test data.}
\label{fig:mort_mgbm_glob_pdp_ice}
\end{figure}

@@

In Figure \ref{fig:mort_mgbm_loc}, local Tree SHAP values are displayed for selected individuals at the 10\textsuperscript{th}, 50\textsuperscript{th}, and 90\textsuperscript{th} percentiles of $g^{\text{MGBM}}(\mathbf{X})$ in the mortgage test data. Each $\phi_j$ value in Figure \ref{fig:mort_mgbm_loc} represents the difference in $g^\text{MGBM}(\mathbf{x^{(i)}})$ and the average of $g^\text{MGBM}(\mathbf{X})$ associated with some input feature $X_j$ \cite{molnar}. Accordingly, the logit of the sum of the Shapley values and the Shapley intercept will be the $g^\text{MGBM}(\mathbf{x})$ prediction in the probability space, for any $\mathbf{x}$.

\begin{figure}[htb]
\centering
\includegraphics[width=16cm]{img/mort_mgbm_loc.png}
\caption{Tree SHAP values for three individuals across selected percentiles of $g^\text{MGBM}(\mathbf{X})$ for the mortgage test data.}
\label{fig:mort_mgbm_loc}
\end{figure}

\noindent The selected individuals show an expected progression of mostly negative Shapley values at the 10\textsuperscript{th} percentile, a mixture of positive and negative Shapley values at the 50\textsuperscript{th} percentile, mostly positive Shapley values the 90\textsuperscript{th} percentile, and with globally important features driving most local model decisions. Deeper significance for Figure \ref{fig:mort_mgbm_loc} lies in the ability of Tree SHAP to accurately and consistently summarize any single $g^\text{MGBM}(\mathbf{x})$ prediction in this manner, which is generally important for enabling logical appeal or override of ML-based decisions, and is specifically important in the context of lending, where applicable regulations often require lenders to provide consumer-specific reasons for denying credit to an individual. In the US, applicable regulations are typically ECOA and FCRA, and the consumer-specific reasons are commonly known as adverse actions codes.

Figure \ref{fig:mort_xnn_glob} displays global feature importance for $g^\text{XNN}$ on the mortgage test data. Deep SHAP values are reported in the probability space, after the application of the logit link function. They are also calculated from the projection layer of $g^\text{XNN}$. Thus, the Deep SHAP values in Figure \ref{fig:mort_xnn_glob} are the estimated average absolute impact of each $X_j$ in the projection layer and probability space of $g^\text{XNN}$ for the mortgage test data. $g^\text{XNN}$ distributes importance more evenly across business drivers and puts stronger emphasis on the no introductory rate period flag feature than does $g^\text{MGBM}$. Like $g^\text{MGBM}$, $g^\text{XNN}$ puts little emphasis on the other flag features.

\begin{figure}[htb]
\centering
\includegraphics[width=9cm]{img/mort_xnn_glob.png}
\caption{Global mean absolute Deep SHAP feature importance for $g^\text{XNN}(\mathbf{X})$ on the mortgage test data.}
\label{fig:mort_xnn_glob}
\end{figure}

@@

Figure \ref{fig:blueprint} displays a holistic approach to ML model training, assessment, deployment meant to decrease discrimination, inaccuracy, privacy, and security risks for high-stakes, human-centered, or regulated ML applications.\footnote{See: \href{https://github.com/jphall663/hc_ml}{Toward Responsible Machine Learning} for details regarding Figure \ref{fig:blueprint}.} While all the methods mentioned in Figure \ref{fig:blueprint} play an important role in increasing human trust and understanding of ML, a few pertinent references and Python resources are highlighted below as further reading to augment this this text's focus on certain interpretable models, post-hoc explanation, and discrimination testing techniques.

@@

\noindent See \href{https://github.com/jphall663/awesome-machine-learning-interpretability}{Awesome Machine Learning Interpretability} for a longer, community-curated metalist of related software packages and resources.

@@

A significant body of research has emerged around exploring and fixing algorithmic discrimination \cite{friedler2019comparative}. Broadly, methodologies can be placed into two groups: more traditional methods that mitigate discrimination by searching across possible algorithmic and feature specifications, and more recently developed approaches that change the algorithms or input data themselves in order to mitigate disparities. Many approaches that have been developed in the last 5--7 years focus on altering the algorithm itself, preprocessing training data, or post-processing predictions in order to diminish class-control correlations or dependencies. Whether these more recent methods are suitable for a particular use case depends on the legal environment in which the model is deployed and on the use case itself. For comments on why these techniques could result in regulatory non-compliance in certain scenarios, see Appendix \ref{a_sec:comp}.
